Categories of Optics

### 2018-06-13 paper.tex
More fiddling

\newcommand{\fget}{\textsc{Get}}
\newcommand{\fput}{\textsc{Put}}
\newcommand{\fmodify}{\textsc{Modify}}
\newcommand{\freview}{\textsc{Review}}
\newcommand{\fmatching}{\textsc{Matching}}
\newcommand{\funzip}{\textsc{Unzip}}

@@

@@

\todo{Contributions:
  \begin{itemize}
  \item A definition of optics general enough to encompass lenses, setters, traversals, and others?
  \item A universal property satisfied by the category of optics in the monoidal case
  \item A satisfying categorical description of lens laws
  \item An attempt to isolate exactly what $\Set$-like properties are being used in various places
  \item A couple of new avenues for generalising optics
  \end{itemize}
}

@@

We begin by constructing the category of optics in a symmetric monoidal category $\C$. This category was first defined in \cite[Section 6]{Doubles} to investigate the relationship between Tambara modules and the center of a monoidal category. There it is called the \emph{double} $\Double_l \C$ of a monoidal category, and our definition is identical, other than flipping the direction of the morphisms to match the existing work on lenses, and restricting our attention to the unenriched case.

@@

For a fixed $M \in \M$, there is a map $\C(S, M A) \times \C(M A, S) \to \C(S, S)$ given by composition. This induces a map $\outside : \Optic_\M(S, A) \to \C(S, S)$, given equationally by $\outside(\rep{l}{r}) = l ; r$

There are also maps two maps
\begin{align*}
  \once, \twice : \Optic_\M(S, A) \to \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S)
\end{align*}
defined by
\begin{align*}
  \once(\rep{l}{r}) &= \repthree{ l}{\id_{MA}}{r} \\
  \twice(\rep{l}{r} \rangle) &= \repthree{l}{r;l}{r}
\end{align*}

@@

We will see that in all the standard examples of optics, this specialises to the expected laws for that type of optic.

@@

  For the second, we must show
  \[ \repthree{ l;(Ml')}{(Mr'); r;l;(Ml')}{(Mr') ; r } = \rep{l;(Ml')}{\id_{MNA}}{(Mr') ; r }. \]

@@

The general pattern is as follows. Once we choose a particular monoidal action $\M \to [\C, \C]$, we find an isomorphism between the set of optics $(S, S') \hto (A, A')$ and a set $\mathbb{O}((S, S'), (A, A'))$ that is easier to describe. Ideally, we then find a similar description $\mathbb{D}(S, A)$ for the set \[ \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S)\] appearing in the definition of a lawful optic. We can then ``read off'' what conditions on an element of $\mathbb{O}((S, S), (A, A))$ are needed to ensure that the corresponding element of $\Optic_\M(S, A)$ is lawful.

\todo{As we get higher up in the hierarchy, the `ad-hoc' laws get closer to being the same as the coend laws}

@@

\todo{This last proposition should probably be deleted now}

In favourable conditions, we can upgrade this result to show that the existence of a lens $S \hto A$ implies that $S \cong C \times A$ for some $C$. This is a generalised from the argument given in~\cite[Corollary 13]{AlgebrasAndUpdateStrategies} for $\Set$.

@@

For an optic, the corresponding element of $\mathbb{O}((S, S'), (A, A'))$ will be called $\funzip$. The above isomorphism sends $\funzip$ to the element $\rep{\funzip }{\ev_{A', S'} }$, where $\ev_{A', S'} : \homC(A',S') \otimes A' \to S'$ is the evaluation map. In the other direction, given $\rep{l}{r }$ we have $\funzip = (\tilde r \otimes A)l$ where $\tilde r : M \to \homC(A', S')$ denotes the transpose of $r$.

\begin{proposition}
  A linear lens $p : S \hto A$ is lawful iff the following two laws for $\funzip$ hold:
  \begin{align*}
    \ev_{A, S} \; \funzip &= \id_S && \textsc{(Rezip)} \\
    (\coev_{\homC(A, S), A} \otimes A)\funzip &= ((\funzip \circ -) \otimes A)\funzip && \textsc{(ZipZip)}
  \end{align*}
  where \[ \funzip \circ - : \homC(A, S) \to \homC(A, \homC(A, S) \otimes A) \] denotes the internal composition. \todo{I obviously need better names for the laws}
\end{proposition}
\begin{proof}
  We have an isomorphism:
  \begin{align*}
    & \int^{M_1, M_2 \in \C} \C(S, M_1 \otimes A) \times \C(M_1 \otimes A, M_2 \otimes A) \times \C(M_2 \otimes A, S) \\
    &\cong \int^{M_1, M_2 \in \C} \C(S, M_1 \otimes A) \times \C(M_1, \homC(A, M_2 \otimes A)) \times \C(M_2, \homC(A, S)) \\
    &\cong \int^{M_1 \in \C} \C(S, M_1 \otimes A) \times \C(M_1, \homC(A, \homC(A, S) \otimes A)) \\
    &\cong \C(S, \homC(A, \homC(A, S) \otimes A) \otimes A)
  \end{align*}
  which evaluated on an element $\repthree{l}{c}{r }$ is given by
  \begin{align*}
    \Phi(\repthree{l}{c}{r}) &= (\widetilde{(\tilde{r} \otimes A)c} \otimes A)l
  \end{align*}
  Now suppose we are given an optic $p$ with corresponding element of $\mathbb{O}(S, A)$ is $\funzip$. Then \[\outside(\rep{\funzip }{\ev_{A, S} }) = \ev_{A, S} \; \funzip,\] giving the first law.

  Now we evaluate $\Phi(\once(p))$ and $\Phi(\twice(p))$.
  \begin{align*}
    \Phi(\once(\rep{\funzip }{\ev_{A, S} }))
    &= \Phi(\repthree{\funzip }{\id_{\homC(A,S) \otimes A}}{\ev_{A, S} }) \\
    &= (\widetilde{(\tilde{\ev_{A, S}} \otimes A)\id_{\homC(A,S) \otimes A}} \otimes A)\funzip \\
    &= (\coev_{\homC(A, S), A} \otimes A)\funzip \\
    \Phi(\twice(\rep{\funzip }{\ev_{A, S} }))
    &= \Phi(\repthree{\funzip }{\funzip \; \ev_{A, S}}{\ev_{A, S} }) \\
    &= (\widetilde{(\tilde{\ev_{A, S}} \otimes A)(\funzip \; \ev_{A, S})} \otimes A)\funzip \\
    &= (\widetilde{(\funzip \; \ev_{A, S})} \otimes A)\funzip \\
    &= ((\funzip \circ -) \otimes A)\funzip
  \end{align*}
  \todo{Fix those overtildes}
  So indeed the optic is lawful iff $(\coev_{\homC(A, S), A} \otimes A)\funzip = ((\funzip \circ -) \otimes A)\funzip$.
\end{proof}

Suppose $\C$ is cartesian closed and has finite coproducts. Then $\C$ acts on itself via $M \cdot A = (M \sqcup 1) \times A$.
\begin{align*}
  \Optic((S, S'), (A, A'))
  &= \int^{M \in \C} \C(S, M \cdot A) \times \C(M \cdot A', S') \\
  &= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \sqcup 1) \times A', S') \\
  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \times A') \sqcup A', S') \\
  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M \times A', S') \times \C(A', S') \\
  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M, \homC(A', S')) \times \C(A', S') \\
  &\cong \C(S, (\homC(A', S') \sqcup 1) \times A) \times \C(A', S') \\
  &\cong \C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S') \\
\end{align*}

\todo{Is it worth working the laws out?}

@@

\begin{lemma}
  Suppose $\C$ is cartesian closed and $F : \C \to \C$ is strong with respect to $\times$. Then for any objects $A, B \in \C$, \[\C(A, FB) \cong \Strong_\C(\times)(A \times \homC(B, -), F)\]
\end{lemma}
\begin{proof}
  We construct the isomorphism $\Phi: C(A, FB) \cong \Strong_\C(\times)(A \times \homC(B, -), F)$ and its inverse $\Psi$.
  Given a $f \in \C(A, FB)$, let $\Phi(f)$ be the natural transformation such that the component at $C$ is the composite
  \begin{align*}
    A \times \homC(B, C) \to FB \times \homC(B, C) \to F(B \times \homC(B, C)) \to FC
  \end{align*}
  I.e.,
  \begin{align*}
    A \times \homC(B, C) &\to FC \\
    (a, g) &\mapsto F(\lambda x. \lambda h. h x)(\theta_{\homC(B, C), B}(g, f a))
  \end{align*}
  % \begin{align*}
      %       A \times \homC(B, C) \xrightarrow{f \times \id} FB \times \homC(B, C) \xrightarrow{\theta_{\homC(B, C), B}}  F(B \times \homC(B, C)) \xrightarrow{F\epsilon} FC
      %     \end{align*}
  where $\theta$ is the strength for $F$.
  \todo{Check this is a strong natural transformation}

  Conversely, given a strong natural transformation $\eta : A \times \homC(B, -) \Rightarrow F$, the component at $B$ is a function $\eta_B : A \times \homC(B, B) \to FB$; we define $\Psi(\eta) : A \to FB$ by $\Psi(\eta)(a) = \eta_B(a, \id_B)$

@@

  Now, verifying these are inverse:
  \begin{align*}
    \Phi(\Psi(\eta))_C (a,g)
    &= F(\lambda x. \lambda h. h x)(\theta_{\homC(B, C), B}(g, \Psi(\eta)(a))) \\
    &= F(\lambda x. \lambda h. h x)(\theta_{\homC(B, C), B}(g, \eta_B(a, \id_B))) \\
    &= F(\lambda x. \lambda h. h x)(\theta_{\homC(B, C), B}(\id, \eta_B)(g, (a, \id_B))) \\
    &= F(\lambda x. \lambda h. h x)(\eta_{\homC(B, C) \times B} s_{\homC(B, C), B}(g, a, \id_B)) && \text{($\theta$ is a strong transformation)}\\
    &= F(\lambda x. \lambda h. h x)(\eta_{\homC(B, C) \times B} (a, \lambda d. (g, \id_B d)) ) && \text{(definition of $s$)}\\
    &= F(\lambda x. \lambda h. h x)(\eta_{\homC(B, C) \times B} (a, \lambda d. (g, d)) ) \\
    &= \eta_{C}(\lambda (a, f). (a, \lambda b. \mathsf{fst}(f b)\mathsf{snd}(f b)) (a, \lambda d. (g, d)) && \text{(naturality of $\eta$)}\\
    &= \eta_{C}(a, \lambda b. g b) \\
    &= \eta_{C}(a, g)
  \end{align*}
  And in the other direction:
  \begin{align*}
    \Psi(\Phi(f))(a)
    &= \Phi(f)_B(a, \id_B) \\
    &= F(\lambda x. \lambda h. h x)(\theta_{\homC(B, B), B}(\id_B, f a))\\
    &= F(\lambda x. \lambda h. h x)(F(\lambda b. (\id_B, b)))\theta_{1, B}(f a) && \text{(naturality of $\theta$)}\\
    &= F(\lambda b. \id_B b)\theta_{1, B}(f a)\\
    &= \theta_{1, B}(f a) && \text{(strength of $\theta$)}\\
    &= f a
  \end{align*}
\end{proof}

This is a generalisation of~\cite[Proposition 2.2]{SecondOrderFunctionals}, isolating the properties of $\Set$ that are used. We can now construct setters:

@@

Due to the above lemma, we can eliminate the coend from the definition:
\begin{align*}
  \Setter_\C((S, S'), (A, A')) &= \int^{F \in \Strong(\C)} \C(S, FA) \times \C(FA', S') \\
                               &\cong \int^{F \in \Strong(\C)} \Strong_\C(\times)(S \times \homC(A, -), F)  \times \C(FA', S') \\
                               &\cong \C(S \times \homC(A, A'), S') \\
                               &\cong \C(\homC(A, A'), \homC(S,S'))
\end{align*}

\todo{laws}

\todo{This previous section might also work just with monoidal closed?}

@@

A folklore variant of a prism is the \emph{affine traversal}, which allows access to a target that may or may not be present. These are described in~\cite{AffineTraversalPost}, and are there hypothesised to correspond to pairs of maps $\C(S, A \sqcup S') \times \C(S\times A', S')$. This is not precisely correct, but very close:

@@

This set includes into $\C(S, A \sqcup S') \times \C(S\times A', S')$, but the two are not equal. In other words, there are fewer affine traversals than expected.

@@

Given such a polymorphic optic in $[\D, \C]$, we can always `monomorphise' to obtain an ordinary optic in $\C$.
\begin{proposition}
  There is a functor
  \begin{align*}
    \mathsf{mono} : \D \times \Optic_{[\D, \C]} \to \Optic_\C
  \end{align*}
  that sends an object $D \in \D$ and optic $\rep{l}{r } : F \hto G$ in $\Optic_{[\D, \C]}$ to the optic $\rep{l_D}{r_D } : FD \hto GD$ in $\Optic_\C$.
\end{proposition}
\begin{proof}
  \todo{todo}
\end{proof}

Lawful optics in $\Optic_{[\D, \C]}$ are sent to lawful optics in $\Optic_\C$,  so we may restrict the domain to obtain a functor:
\begin{align*}
  \mathsf{mono} : \D \times \Lawful_{[\D, \C]} \to \Lawful_\C
\end{align*}

\todo{more to say?}

@@

>>>>

\newcommand{\fget}{\textsc{Get}}
\newcommand{\fput}{\textsc{Put}}
\newcommand{\fmodify}{\textsc{Modify}}
\newcommand{\freview}{\textsc{Review}}
\newcommand{\fmatching}{\textsc{Matching}}
\newcommand{\funzip}{\textsc{Unzip}}
\newcommand{\fover}{\textsc{Over}}

@@

% Credit: Sam Buss
\newbox\gnBoxA
\newdimen\gnCornerHgt
\setbox\gnBoxA=\hbox{$\ulcorner$}
\global\gnCornerHgt=\ht\gnBoxA
\newdimen\gnArgHgt
\def\nameof #1{%
    \setbox\gnBoxA=\hbox{$#1$}%
    \gnArgHgt=\ht\gnBoxA%
    \ifnum     \gnArgHgt<\gnCornerHgt \gnArgHgt=0pt%
    \else \advance \gnArgHgt by -\gnCornerHgt%
    \fi \raise\gnArgHgt\hbox{$\ulcorner$} \box\gnBoxA %
    \raise\gnArgHgt\hbox{$\urcorner$}}

@@

\todo{Contributions:
  \begin{itemize}
  \item A definition of optics general enough to encompass lenses, setters, traversals, and others?
  \item A definition of lawful optic that specialises in the correct way and allows us to easily derive ad-hoc laws for new kids of optic
  \item A universal property satisfied by the category of optics in the monoidal case
  \item A satisfying categorical description of lens laws
  \item An attempt to isolate exactly what $\Set$-like properties are being used in various places
  \item New avenues for generalising optics
  \end{itemize}
}

@@

We begin by constructing the category of optics in a symmetric monoidal category $\C$. This category was first defined in \cite[Section 6]{Doubles} to investigate the relationship between Tambara modules and the center of a monoidal category. There it is called the \emph{double} of a monoidal category, and our definition is identical, other than flipping the direction of the morphisms to match the existing work on lenses and restricting our attention to the unenriched case.

@@

Using the universal property of the coend, we define three maps:
\begin{align*}
  \outside &: \Optic_\M(S, A) \to \C(S, S) \\
  \once, \twice &: \Optic_\M(S, A) \to \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S)
\end{align*}
by
\begin{align*}
  \outside(\rep{l}{r}) &= l;r \\
  \once(\rep{l}{r}) &= \repthree{l}{\id_{MA}}{r} \\
  \twice(\rep{l}{r}) &= \repthree{l}{r;l}{r}
\end{align*}

@@

We will see that in all the common examples of optics, this specialises to the expected laws for that variant of optic.

\begin{remark}
The above definition of lawful does not match neatly with the constant complement idea, which might instead require that $r;l = \id_{MA}$ for some $M$. I argue that $\once(p) = \twice(p)$ is the correct condition for a few reasons.

The main reason is that it corresponds precisely to the ad-hoc lens laws with no additional requirements, whereas finding a complement for a lens requires further conditions on the category $\C$.

Secondly, even if $r;l = \id_{MA}$ holds for a particular representative $\langle l, r \rangle$, the same is not likely to be true for a different representative.

Finally, we will later see an interpretation of optics as morphisms between certain comonoid objects. Lawfulness corresponds exactly to a morphism being a comonoid morphism.
\end{remark}

@@

  For the second, we must show
  \[ \repthree{ l;(Ml')}{(Mr'); r;l;(Ml')}{(Mr') ; r } = \repthree{l;(Ml')}{\id_{MNA}}{(Mr') ; r }. \]

@@

The general pattern is as follows. Once we choose a particular monoidal action $\M \to [\C, \C]$, we find an isomorphism between the set of optics $(S, S') \hto (A, A')$ and a set $\mathbb{O}((S, S'), (A, A'))$ that is easier to describe. Ideally, we then also find a simplified description $\mathbb{D}(S, A)$ for the set \[ \int^{M_1, M_2 \in \M} \C(S, M_1 A) \times \C(M_1 A, M_2 A) \times \C(M_2 A, S)\] appearing in the definition of a lawful optic. We can then ``read off'' what conditions on an element of $\mathbb{O}((S, S), (A, A))$ are needed to ensure that the corresponding element of $\Optic_\M(S, A)$ is lawful. With no offence intended, we will call these more specific conditions the `ad-hoc' laws.

\todo{There is a common pattern in some of these examples, where the evaluation-at-$A$ functor $-A : \M \to \C$ has a left or right adjoint. Is it worth abstracting this out?}

@@

If we ask more of our category $\C$, we can upgrade this result to show that a lens $S \hto A$ implies the existence of a complement $\C$ with $S \cong C \times A$. This is a generalised from the argument given in~\cite[Corollary 13]{AlgebrasAndUpdateStrategies} for $\Set$.

@@

@@

\begin{lemma}
  Suppose $\C$ is cartesian closed and $F : \C \to \C$ is strong with respect to $\times$. Then for any objects $A, B \in \C$, \[\C(A, FB) \cong \Strong_\C(\times)(A \times \homC(B, -), F)\]
naturally in $A$ and $F$. In other words, the functor $(A \mapsto A \times \homC(B, -)) : \C \to \Strong_\C$ is left adjoint to the evaluation-at-$B$ functor $-B : \Strong_\C \to \C$.
\end{lemma}
\begin{proof}
  We construct the isomorphism $P : C(A, FB) \cong \Strong_\C(\times)(A \times \homC(B, -), F)$ and its inverse $Q$.
  Given a $f \in \C(A, FB)$, let $P(f)$ be the natural transformation such that the component at $C$ is the composite
  \begin{align*}
    A \times \homC(B, C) \to FB \times \homC(B, C) \to F(B \times \homC(B, C)) \to FC
  \end{align*}
  I.e.,
  \begin{align*}
    A \times \homC(B, C) &\to FC \\
    (a, g) &\mapsto F(\lambda x. \lambda h. h x)(\theta_{\homC(B, C), B}(g, f a))
  \end{align*}
  % \begin{align*}
      %       A \times \homC(B, C) \xrightarrow{f \times \id} FB \times \homC(B, C) \xrightarrow{\theta_{\homC(B, C), B}}  F(B \times \homC(B, C)) \xrightarrow{F\epsilon} FC
      %     \end{align*}
  where $\theta$ is the strength for $F$.
  \todo{Check this is a strong natural transformation}

  Conversely, given a strong natural transformation $\eta : A \times \homC(B, -) \Rightarrow F$, the component at $B$ is a function $\eta_B : A \times \homC(B, B) \to FB$; we define $Q(\eta) : A \to FB$ by $Q(\eta)(a) = \eta_B(a, \id_B)$

@@

  Now, verifying these are inverse:
  \begin{align*}
    P(Q(\eta))_C (a,g)
    &= F(\lambda x. \lambda h. h x)(\theta_{\homC(B, C), B}(g, \Psi(\eta)(a))) \\
    &= F(\lambda x. \lambda h. h x)(\theta_{\homC(B, C), B}(g, \eta_B(a, \id_B))) \\
    &= F(\lambda x. \lambda h. h x)(\theta_{\homC(B, C), B}(\id, \eta_B)(g, (a, \id_B))) \\
    &= F(\lambda x. \lambda h. h x)(\eta_{\homC(B, C) \times B} s_{\homC(B, C), B}(g, a, \id_B)) && \text{($\theta$ is a strong transformation)}\\
    &= F(\lambda x. \lambda h. h x)(\eta_{\homC(B, C) \times B} (a, \lambda d. (g, \id_B d)) ) && \text{(definition of $s$)}\\
    &= F(\lambda x. \lambda h. h x)(\eta_{\homC(B, C) \times B} (a, \lambda d. (g, d)) ) \\
    &= \eta_{C}(\lambda (a, f). (a, \lambda b. \mathsf{fst}(f b)\mathsf{snd}(f b)) (a, \lambda d. (g, d)) && \text{(naturality of $\eta$)}\\
    &= \eta_{C}(a, \lambda b. g b) \\
    &= \eta_{C}(a, g)
  \end{align*}
  And in the other direction:
  \begin{align*}
    Q(P(f))(a)
    &= P(f)_B(a, \id_B) \\
    &= F(\lambda x. \lambda h. h x)(\theta_{\homC(B, B), B}(\id_B, f a))\\
    &= F(\lambda x. \lambda h. h x)(F(\lambda b. (\id_B, b)))\theta_{1, B}(f a) && \text{(naturality of $\theta$)}\\
    &= F(\lambda b. \id_B b)\theta_{1, B}(f a)\\
    &= \theta_{1, B}(f a) && \text{(strength of $\theta$)}\\
    &= f a
  \end{align*}
\todo{Naturality}
\end{proof}

This is a generalisation of~\cite[Proposition 2.2]{SecondOrderFunctionals}, isolating the properties of $\Set$ that are used. We can finally construct setters:

@@

Due to the above lemma, we can eliminate the coend from the definition:
\begin{align*}
  \Setter_\C((S, S'), (A, A')) &= \int^{F \in \Strong(\C)} \C(S, FA) \times \C(FA', S') \\
                               &\cong \int^{F \in \Strong(\C)} \Strong_\C(\times)(S \times \homC(A, -), F)  \times \C(FA', S') \\
                               &\cong \C(S \times \homC(A, A'), S') \\
                               &\cong \C(\homC(A, A'), \homC(S,S'))
\end{align*}
In the Haskell \lenslib{} library, the map $\homC(A, A) \to \homC(S,S)$ corresponding to a setter is called $\fover$: we think of a setter as allowing us to apply a function $A \to A$ over some parts of $S$.

The laws for setters are a kind of `functoriality':

\begin{proposition}
A setter $p : S \hto A$ is lawful iff 
\begin{align*}
\fover \; \nameof{\id_A} &= \nameof{\id_S} \\
({-} \circ =)(\fover \times \fover) &= \fover ({-} \circ =)
\end{align*}
where  $\circ$ denotes internal composition and $\nameof{\id_A} : 1 \to \homC(A, A)$ is the `name of' $\id_A$, and similarly for $\id_S$.
\end{proposition}
\begin{proof}
We have a chain of isomorphisms
\begin{align*}
&\int^{M_1, M_2 \in \Strong_\C} \C(S, M_1 A) \times \C(M_1 A, M_2  A) \times \C(M_2 A, S) \\
&\cong \int^{M_1, M_2 \in \Strong_\C} \Strong_\C(S \times \homC(A, -), M_1) \times \Strong_\C(M_1 A \times \homC(A, -), M_2) \times \C(M_2 A, S) \\
&\cong \int^{M_2 \in \Strong_\C} \Strong_\C(S \times \homC(A, A) \times \homC(A, -), M_2) \times \C(M_2 A, S) \\
&\cong \C(S \times \homC(A, A) \times \homC(A, A), S) \\
&\cong \C(\homC(A, A) \times \homC(A, A), \homC(S,S))
\end{align*}
\todo{Worth writing out the rest?}
\end{proof}

\todo{This previous section might also work just with monoidal closed? I don't see where cartesianness was used.}

For an optic, the corresponding element of $\mathbb{O}((S, S'), (A, A'))$ will be called $\funzip$. The above isomorphism sends $\funzip$ to the element $\rep{\funzip }{\ev_{A', S'}}$, where $\ev_{A', S'} : \homC(A',S') \otimes A' \to S'$ is the evaluation map. In the other direction, given $\rep{l}{r }$ we have $\funzip = (\tilde r \otimes A)l$ where $\tilde r : M \to \homC(A', S')$ denotes the transpose of $r$.

\begin{proposition}
  A linear lens $p : S \hto A$ is lawful iff the following two laws for $\funzip$ hold:
  \begin{align*}
    \ev_{A, S} \; \funzip &= \id_S && \textsc{(Rezip)} \\
    (\coev_{\homC(A, S), A} \otimes A)\funzip &= ((\funzip \circ -) \otimes A)\funzip && \textsc{(ZipZip)}
  \end{align*}
  where \[ \funzip \circ - : \homC(A, S) \to \homC(A, \homC(A, S) \otimes A) \] denotes internal composition and \[\coev_{\homC(A, S), A} : \homC(A, S) \to \homC(A, \homC(A, S) \otimes A)\] is coevaluation.
  \todo{I obviously need better names for the laws}
\end{proposition}
\begin{proof}
  We have an isomorphism:
  \begin{align*}
    & \int^{M_1, M_2 \in \C} \C(S, M_1 \otimes A) \times \C(M_1 \otimes A, M_2 \otimes A) \times \C(M_2 \otimes A, S) \\
    &\cong \int^{M_1, M_2 \in \C} \C(S, M_1 \otimes A) \times \C(M_1, \homC(A, M_2 \otimes A)) \times \C(M_2, \homC(A, S)) \\
    &\cong \int^{M_1 \in \C} \C(S, M_1 \otimes A) \times \C(M_1, \homC(A, \homC(A, S) \otimes A)) \\
    &\cong \C(S, \homC(A, \homC(A, S) \otimes A) \otimes A)
  \end{align*}
  which evaluated on an element $\repthree{l}{c}{r }$ is given by
  \begin{align*}
    \Phi(\repthree{l}{c}{r}) &= (\widetilde{(\tilde{r} \otimes A)c} \otimes A)l
  \end{align*}
  Now suppose we are given an optic $p$ with corresponding element of $\mathbb{O}(S, A)$ is $\funzip$. Then \[\outside(\rep{\funzip }{\ev_{A, S} }) = \ev_{A, S} \; \funzip,\] giving the first law.

  Now we evaluate $\Phi(\once(p))$ and $\Phi(\twice(p))$.
  \begin{align*}
    \Phi(\once(\rep{\funzip }{\ev_{A, S} }))
    &= \Phi(\repthree{\funzip }{\id_{\homC(A,S) \otimes A}}{\ev_{A, S} }) \\
    &= (\widetilde{(\tilde{\ev_{A, S}} \otimes A)\id_{\homC(A,S) \otimes A}} \otimes A)\funzip \\
    &= (\widetilde{(\id_{\homC(A, S)} \otimes A)} \otimes A)\funzip \\
    &= (\coev_{\homC(A, S), A} \otimes A)\funzip \\
    \Phi(\twice(\rep{\funzip }{\ev_{A, S} }))
    &= \Phi(\repthree{\funzip }{\funzip \; \ev_{A, S}}{\ev_{A, S} }) \\
    &= (\widetilde{(\tilde{\ev_{A, S}} \otimes A)(\funzip \; \ev_{A, S})} \otimes A)\funzip \\
    &= (\widetilde{(\funzip \; \ev_{A, S})} \otimes A)\funzip \\
    &= ((\funzip \circ -) \otimes A)\funzip
  \end{align*}
  \todo{Fix those overtildes}
  So indeed the optic is lawful iff $\ev_{A, S} \; \funzip = \id_S$ and $(\coev_{\homC(A, S), A} \otimes A)\funzip = ((\funzip \circ -) \otimes A)\funzip$.
\end{proof}

Suppose $\C$ is cartesian closed and has finite coproducts. Then $\C$ acts on itself via $M \cdot A = (M \sqcup 1) \times A$.
\begin{align*}
  \Optic((S, S'), (A, A'))
  &= \int^{M \in \C} \C(S, M \cdot A) \times \C(M \cdot A', S') \\
  &= \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \sqcup 1) \times A', S') \\
  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C((M \times A') \sqcup A', S') \\
  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M \times A', S') \times \C(A', S') \\
  &\cong \int^{M \in \C} \C(S, (M \sqcup 1) \times A) \times \C(M, \homC(A', S')) \times \C(A', S') \\
  &\cong \C(S, (\homC(A', S') \sqcup 1) \times A) \times \C(A', S') \\
  &\cong \C(S, \homC(A', S') \sqcup 1) \times \C(S, A) \times \C(A', S')
\end{align*}

\todo{Worth writing out?}

@@

A folklore variant of a prism is the \emph{affine traversal}, which allows access to a target that may or may not be present. These are described in~\cite{AffineTraversalPost}, and it is suggested they correspond to pairs of maps $\C(S, A \sqcup S') \times \C(S\times A', S')$. This is not precisely correct, but very close:

@@

This set includes into $\C(S, A \sqcup S') \times \C(S\times A', S')$, but the two are not equal.

@@

Given such a polymorphic optic in $[\D, \C]$, we can always `monomorphise' to obtain an ordinary optic in $\C$.
\begin{proposition}
  There is a functor
  \begin{align*}
    \mathsf{mono} : \D \times \Optic_{[\D, \C]} \to \Optic_\C
  \end{align*}
  that sends an object $D \in \D$ and optic $\rep{l}{r } : F \hto G$ in $\Optic_{[\D, \C]}$ to the optic $\rep{l_D}{r_D } : FD \hto GD$ in $\Optic_\C$. Further, this preserves lawfulness.
\end{proposition}
\begin{proof}
  \todo{todo}
\end{proof}

@@

\todo{Maybe some of the long equational manipulations could be done by commutative diagram, using a pair of dashes to signal the start and end value?}
